Investigation of thermal reactionomes reveals stress tolerance vs. resistance in

### 2014-08-19 doc/ApTranscriptome_manuscript.Rmd
edits to Intro from SHC

---
title: Investigations of thermal reactionomes reveal stress tolerance vs. resistance in
  warm and cold-climate ant species \\
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the evolution of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. Interestingly, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change to thermal extremes than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress. Namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences within and among species in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. How this variation in thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], among other traits. These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few levels of temperature. However, temperature varies continuously in nature. As a results, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014]. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is. What is needed is a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].

@@

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences. Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes to gain insight into how spatially divergent selection has led to patterns thermally-responsive gene expression in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. Temperature is a potentially strong selective force on ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].

@@

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012].

>>>>

---
title: Investigation of thermal reactionomes reveals stress tolerance vs. resistance in
  warm and cold-climate ant species \\
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the genetic basis of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression chang than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* also shut down more genes in response to thermal extremes than *A. picea*. A subset of these genes were upregulated in *A. picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress. Namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

# Significance Statement

Thermal tolerance is a key parameter underlying species distributions, and varies considerably within species. However, little is known about the genetic basis of thermal tolerance or how it has evolved. In recent years, whole organism gene expression (e.g. transcriptomic) studies have been used to gain greater insight on these questions, but these studies are limited to finding genes with large effects in a few conditions. Here, we present a method to characterize the reaction norms of gene expression across an ecologically-relevant thermal gradient (e.g. the thermal reactionome) for all genes in the transcriptomes of two closely related ant species. We find that a large fraction of the transcriptome is thermally-responsive. The warmer-climate species down-regulates more genes in response to thermal extremes than the colder-climate species. These differences in expression patterns are indicative of divergent strategies in these two species; stress tolerance in the warmer-climate species and stress-resistance in the colder-climate species.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance. This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short term acclimation and adaptation to increased temperatures [@somero2010].

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few environmental levels. However, temperature and other environmental factors vary continuously in nature. As a result, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014]. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is. What is needed is a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].

@@

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], and then compare the thermal reactionomes of these species to gain insight into how spatially divergent selection has led to changes in thermally-responsive gene expression.

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010]. Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].

@@
